The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-07-02 paper/referee-reply.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{graphicx}

\title{Response to the first report on the article draft \emph{The
    Impact of Gas Bulk Rotation on the Lyman-$\alpha$ line.}} 
\author{Juan N. Garavito-Camargo, Jaime E. Forero-Romero, Mark Disjkstra}
\date{\today}

\begin{document}

\maketitle
We thank the referee for a detailed report. It prompted us to explore
a very important aspect of our results that we had overlooked. Namely,
the dependence of the line morphology as a function of viewing
angle. In the revision process we have also corrected a mistake in the
initialization procedure for the MC calculations. This also allowed us
to derive an analytical solution that reproduced the main features of
the MC results.

The improvement in these three aspects (line anisotropy, photon initialization,
analytical solution) has led us to rewrite the Results and Discussion
section. We have also completeley dropped sections (e.g. the asymetrical emission
part) that obfuscate the new main results of the paper.

Given the substantial changes from the first version, 
we kindly ask the referee to revise again in detail the new manuscript.

@@

The result is found in both codes.

\subsection*{Fig4 and 6}

{\bf This is a surprising result that the number of scatterings
  (escape fraction) stays constant as the rotation velocity increases,
  for a central source, whereas the global spectral shape is
  altered. Did you try with higher/extreme values of Vmax (=1000 km/s,
  even if not physically motivated) ? Do you believe that the number
  of scatterings decreases with very high values, or that it is
  independant of the rotation velocity ? Is the escape fraction from a
  dusty rotating cloud with central source independant of the rotation
  velocity ?}

This results is not surprising if one considers that the scatterings
proceed as in the static case. In the co-rotating frame everything
happens exactly as in the estatic case.

We have also veryfied that for high rotational velocities of 1000 km/s
the results for the escape fraction, number of scatterings and integrated total
flux remain constant.

@@

\subsection*{Fig5}

{\bf This is a very nice figure ! Looking at the top right panel, with
  its “photosphere”, I’m surprised that the distribution of Nscatt is
  bipolar, I would have expected a smooth transition between the 2
  regimes. Do you have a physical explanation why photons escape after
  either (less than 10) or (more than 1000) scatterings, and none
  escape with 100 scatt ? To test the ’photosphere’ assumption, you
  could also do the radiation transfer in a cloud with the
  distribution of sources only in an inner sphere, and another
  experienment with sources only in an outer shell. If I understood
  corectly, you would expect thant the bottom spot on the top right
  panel is made of photons from the outer shell, and the uppper spot
  from photons emitted inside the cloud.}

Actully this plot is incorrect due to the wrong initialization of the
photons in the homogeneous case. In that situation the frequency of
the photons must be changed according to the dot product between the
bulk velocity and propagation direction. This is mentioned in section
2.2.

With this changes we find that the total number of scatterings does
not change with rotational velocity.

@@

\subsection*{End of paragraph 3.3}

{\bf One sentence is not clear : we see that the escape fraction does
  not increase significantly from $\tau$ = 105 to $\tau =
  10^{6}$. This counter-intuitive result.... It sounds like you were
  expecting a strong increase... A decrease is expected from $\tau$ =
  105 to $\tau$ = 106, not an increase, but indeed on the graphe we
  can see an unexpected (small) increase. I do not understand the
  explanation for this behaviour.}

We have included the following paragraphs in the paper:

\begin{quote}
The naive interpretation of the analytic solution does not seem to
hold for photons emitted far from the sphere's center. 
We suggest that increasing $\tau_{\rm H}$ from $10^{5}$ to $10^{6}$ causes a
transition from the 'optically thick' to  the 'extremely optically
thick' regime for a noticeable fraction of the photons in the
homogeneous source distribution.

In the optically thick regime, lya  photons can escape in
'single flight' which corresponds to a scenario in which the
photon resonantly scatters $10^4-10^5$ times until it is scattered
into the wing of the line ($x\sim 3-4$). 
At these frequencies the medium is optically thin, and the photons can
escape efficiently in a single flight. 
In contrast, in an extremely optically thick medium lya
  photons escape in a `single excursion' (Adams 72). 
Here, photons that are scattered into the wing of the line escape from
the medium in a sequence of wing scattering events. 
In both cases, lya photons resonantly scatter $10^4$-$10^5$ times. 
Because we keep our clouds the same size, the mean free path of Lya
photons that scatter resonantly is 10 times larger for the case
$\tau_{\rm H}=10^5$ than for $\tau_{\rm H}=10^6$. 
If we compute the average distance $D$ travelled by lya
photons through a medium of size $R$ as a function of line center
optical depth $\tau_{\rm H}$, then we find that during the transition
from optically thick to extremely optically thick the mean traversed
distance $D$ actually decreases slightly.  
This decrease is unique to this transition region, and $D$ generally
increases with $\tau_{\rm H}$ at other values of $\tau_{\rm H}$.   
\end{quote}

@@

\subsection*{Fig8}

{\bf Referred to as Figure 7 in the text (paragraph 3.4). To my mind,
  this figure illustrates the main result of your study, it has to be
  more demonstrative. On this figure, the numerical noise seems very
  big compare to the small number of bins, bigger than on the spectra
  from Fig1, how is it possible ? Could you re-do this figure with
  more photons, maybe with more bins ? Could you plot an histogram
  instead of broken lines, to ease the reading ? In the text, you
  write anisotropy induced by rotation is at the 3$\%$ level, how do
  to estimate this number? I think that the noise is at $3\%$ level,
  and the distribution is compatible with ’flat’. In the corresponding
  text, you first say that $F(\mu)$ does not depend on Vmax. Again, if
  real, this is an important result, that you must advertise
  more. Please show it on this Fig, by comparing F($\mu$) without rotation
  (in red, for example) to the other distributions that you get with
  different values of Vmax. Then, you mention that for high optical
  depth values, $F(\mu)$ (not the variations of $\mu$) can vary up to
  $15\%$. First, what do you call high optical depth ? your highest
  setup is $\tau$= $10^7$, not extremely high for lyman alpha ”standards’,
  corresponding to NHI $\sim$10$^{20}$ cm$^{-2}$. So, if you see a trend with
  optical depth, I encourage you to check higher optical depth
  regimes, corresponding better to galaxy scales. If you can probe
  that anisotropy induced by rotation is rising with optical depth,
  this would be an interesting result. If higher optical depth regimes
  lead to anisotropic escape, could you check the effet of anisotropy
  on the lya spectral shape?}

We have prepared Figure 5 that shows the $F(\mu)$ distribution for
different velocities and optical depths. Indeed, the distributions are
consistent with flat (within the noise levels) for all the models
studied in our paper.

We did not try models with higher optical depths due to the
limitation in available computing time to perform the
calculations. However, the analytical solution provides sufficient
information to quantify the anisotropy level.

### 2014-07-04 paper/ms.tex
minor updated
fixed typos

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{f1.pdf}
\end{center}
\caption{Geometry of the gas distribution. The angular velocity vector
  is parallel to the unit vector $\hat{k}$. In order to describe the
  departures from spherical symmetry we use the polar angle $\theta$
  formed by the direction of the outgoing photons with respecto to the
  $z$-axis. We define define the variable $\mu\equiv\cos\theta$ to
  report to present our results. Computing the spectra for photons in
  a   narrow range of $\mu$ is equivalent to having a line-of-sight
  oriented in that direction.  
    \label{fig:geometry}}  
\end{figure}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f2.pdf}
\end{center}
\caption{
2D histogram showing the number of photons that escape with frecuency
$x$ forming an angle $\theta$ (parameterized as $|\cos\theta|$) with the
rotation axis. 
The rotational velocity ($0,100,200,300$\kms) increases from left to
right and the optical depth ($10^5$, $10^6$, $10^7$) from top to
bottom. 
The \ly photons are initialized at the center of the sphere. 
Two main results can be read from this figure.
First, the line morphology depends on the viewing angle. 
Second, the line can become single peaked for high rotational
velocities.
\label{fig:CentralSpec} }   
\end{figure*}

@@

The initialization process for the \ly photons especifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the gas reference frame and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The central result of this paper are summarized in Fig.s
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parameterized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have veryfied that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Fig.s we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximu, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

If the viewing angle is aligned witht the rotation axis, $|\mu|\sim
1$, the \ly line keeps in most of the cases a double peak with minor
changes as the rotational velocity starts to increase. 
However, for a line of sight perpendicular to the rotation axis,
$|\mu|\sim 0$, the impact of  rotation is larger.  
The double peak readily transforms into a single peak.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{f5.pdf}
\end{center}
\caption{Integrated flux distribution as a function of the
  viewing angle as parameterized by $\mu$. Continous (dashed)
  correspond to central (homogeneous) source distribution.
  The models correspond to an optical depth of $\tau_{\rm
    H}=10^5$ and rotational velocities of $100$\kms, $200$\kms and
  $300$\kms. The distributions are flat in the range of models probed
  in this paper, meaning that the integrated flux for all viewing
  angles is the same.
\label{fig:muhisto}} 
\end{figure}

@@

There is a key result of our simulatios that allows us to build an
analytical description for the outgoing spectra.
It is the independence of the following three quantities with the rotational
velocity and the viewing angle: integrated flux, average number of
scatterings and escape fraction.

One way to understand this is that a rotating sphere
is practically identical from a static one. 
In the frame of the rotating gas cloud all atoms are in fact
stationary with respect to each other, making the scattering process
proceed identical as in the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each othe and the scattering process
proceeds identical as in the static case.

>>>>

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{f1.pdf}
\end{center}
\caption{Geometry of the gas distribution. The angular velocity vector
  is parallel to the unit vector $\hat{k}$. In order to describe the
  departures from spherical symmetry we use the polar angle $\theta$
  formed by the direction of the outgoing photons with respect to the
  $z$-axis. We define define the variable $\mu\equiv\cos\theta$ to
  report to present our results. Computing the spectra for photons in
  a   narrow range of $\mu$ is equivalent to having a line-of-sight
  oriented in that direction.  
    \label{fig:geometry}}  
\end{figure}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f2.pdf}
\end{center}
\caption{
2D histogram showing the number of photons that escape with frequency
$x$ forming an angle $\theta$ (parametrized as $|\cos\theta|$) with the
rotation axis. 
The rotational velocity ($0,100,200,300$\kms) increases from left to
right and the optical depth ($10^5$, $10^6$, $10^7$) from top to
bottom. 
The \ly photons are initialized at the center of the sphere. 
Two main results can be read from this figure.
First, the line morphology depends on the viewing angle. 
Second, the line can become single peaked for high rotational
velocities.
\label{fig:CentralSpec} }   
\end{figure*}

@@

The initialization process for the \ly photons specifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the gas reference frame and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

@@

The central result of this paper are summarized in Fig.s
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. 
They show 2D histograms of the escape frequency $x$ and outgoing angle
$\theta$ parametrized by $|\mu|$. 
Taking into account only photons photons around a value
of $|\mu|$ gives us the emission detected by an observer located at an
angle $\theta$ with respect to the rotation axis. 
We have verified that the solutions are indeed symmetric with respect
to $\mu=0$.

From these Fig.s we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  
In the next subsections we quantify the morphology changes with with
velocity, optical depth and viewing angle.  
We characterize the line morphology by its total intensity, the full
width at half maximum, (FWHM) and the location of the peak maxima. 
In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. 
For the models where dust is included we measure the escape fraction
as a function of rotational velocity.

@@

If the viewing angle is aligned with the rotation axis, $|\mu|\sim
1$, the \ly line keeps in most of the cases a double peak with minor
changes as the rotational velocity starts to increase. 
However, for a line of sight perpendicular to the rotation axis,
$|\mu|\sim 0$, the impact of  rotation is larger.  
The double peak readily transforms into a single peak.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{f5.pdf}
\end{center}
\caption{Integrated flux distribution as a function of the
  viewing angle as parametrized by $\mu$. Continuous (dashed)
  correspond to central (homogeneous) source distribution.
  The models correspond to an optical depth of $\tau_{\rm
    H}=10^5$ and rotational velocities of $100$\kms, $200$\kms and
  $300$\kms. The distributions are flat in the range of models probed
  in this paper, meaning that the integrated flux for all viewing
  angles is the same.
\label{fig:muhisto}} 
\end{figure}

@@

There is a key result of our simulations that allows us to build an
analytical description for the outgoing spectra.
It is the independence of the following three quantities with the rotational
velocity and the viewing angle: integrated flux, average number of
scatterings and escape fraction.

One way to understand this is that a rotating sphere
is practically identical from a static one. 
In the frame of the rotating gas cloud all atoms are in fact
stationary with respect to each other, making the scattering process
proceed identical as in the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each other and the scattering process
proceeds identical as in the static case.
